---
title: "Exploratory Data Analysis of Enron Dataset"
author: "Adam McCarthy"
date: "7/30/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pressure, echo=FALSE}
enron <- read.csv(file = "enron_data.csv")
```

The dataset is the final data set provided by Udacity for the intro
to machine learning course.

It includes a combination of financial data and variables created from
email data.

There are 22 variables.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(ggthemes)
library(scales)
library(extRemes)
library(sm)
library(gridExtra)
library(GGally)
```

#### Descriptive Statistics
```{r echo=FALSE, Summary}
summary(enron)
```

#### Uni-variate analysis
```{r echo=FALSE, Multiplot_function}
# Entirely taken from:
# http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots == 1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```


```{r echo=FALSE, U2, fig.height = 7.5, fig.width = 5, fig.align="center"}
small_hist <- function(data_v, name, scale, df=enron) {
plt <- ggplot(data = df, aes(data_v)) + 
  geom_histogram(bins = 50,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale,
    y = "Count"
    )

return(plt)
}

plt1 <- small_hist(enron$bonus, "Bonus", "$")
plt2 <- small_hist(enron$deferral_payments, "Deferral payments", "$")
plt3 <- small_hist(enron$deferred_income, "Deferred income", "$")
plt4 <- small_hist(enron$exercised_stock_options, "Exercised stock options", "$")
plt5 <- small_hist(enron$expenses, "Expenses", "$")
plt6 <- small_hist(enron$loan_advances, "Loan advances", "$")
plt7 <- small_hist(enron$long_term_incentive, "Long term incentive", "$")
plt8 <- small_hist(enron$other, "Other", "$")
plt9 <- small_hist(enron$restricted_stock, "Restricted stock", "$")
plt10 <- small_hist(enron$restricted_stock_deferred, "Restricted stock deffered", "$")
plt11 <- small_hist(enron$salary, "Salary", "$")
plt12 <- small_hist(enron$total_payments, "Total payments", "$")
plt13 <- small_hist(enron$total_stock_value, "Total stock value", "$")

# Multiplot used when grid.arrange does not work (e.g. not a ggplot)
multiplot(plt1, plt2, plt3, plt4, plt5, 
          plt6, plt7, plt8, plt9, plt10, 
          plt11, plt12, plt13, cols = 3)
```

The histograms of all of the financial variables highlight a variety of
distributions present in the dataset. Many of the variables have missing
values and have these dropped for the plot.

There are no normal distributions, nearly all have skewewd distributions.

```{R echo=FALSE, Log_hist, fig.height = 3, fig.width = 5}
plt1t <-  plt1 + scale_x_log10()
plt2t <-  plt2 + scale_x_log10()
plt3t <-  plt3 + scale_x_sqrt()
plt4t <-  plt4 + scale_x_log10()
plt5t <-  plt5 + scale_x_log10()
plt6t <-  plt6 + scale_x_log10()
plt7t <-  plt7 + scale_x_sqrt()
plt8t <-  plt8 + scale_x_log10()
plt9t <-  plt9 + scale_x_log10()
plt10t <-  plt10 + scale_x_sqrt()
plt11t <-  plt11 + scale_x_log10()
plt12t <-  plt12 + scale_x_sqrt()
plt13t <-  plt13 + scale_x_sqrt()

multiplot(plt1t, plt2t, plt3t, plt4t, plt5t, 
          plt6t, plt7t, plt8t, plt9t, plt10t, 
          plt11t, plt12t, plt13t, cols = 3)
```

Data transforms are applied to better represent the distribution of values.
Either log10 or sqrt are applied. This gives a better visualisation of the
over dispersed variables.

```{r echo=FALSE, U3, fig.height = 3, fig.width = 5, fig.align="center"}
plt14 <- small_hist(enron$from_messages, "From messages", "count")
plt15 <- small_hist(enron$from_poi_to_this_person, "From POI to this person", "count")
plt16 <- small_hist(enron$from_this_person_to_poi, "From this person to POI", "count")
plt17 <- small_hist(enron$shared_receipt_with_poi, "Shared receipt with POI", "count")
plt18 <- small_hist(enron$to_messages, "To message", "count")

# Multiplot used when grid.arrange does not work (e.g. not a ggplot)
multiplot(plt14, plt15, plt16, plt17, plt18, cols = 3)
```

The variables sourced from emails also show highly skewed variables.

```{R echo=FALSE, Log_hist, fig.height = 3, fig.width = 5}
plt20 <- ggplot(aes(x = enron$from_messages), data = enron) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "From messages"
    ) +
  labs(
    x = "Number log10",
    y = "Count"
    ) + scale_x_log10()

plt21 <- ggplot(aes(x = enron$from_this_person_to_poi), data = enron) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "From this person to POI"
    ) +
  labs(
    x = "Number log10",
    y = "Count"
    ) + scale_x_log10()

plt22 <- ggplot(aes(x = enron$to_messages), data = enron) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "To messages"
    ) +
  labs(
    x = "Number log10",
    y = "Count"
    ) + scale_x_log10()

plt23 <- ggplot(aes(x = enron$from_poi_to_this_person), data = enron) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "From POI to this person"
    ) +
  labs(
    x = "Number log10",
    y = "Count"
    ) + scale_x_log10()

plt24 <- ggplot(aes(x = enron$shared_receipt_with_poi), data = enron) +
  geom_histogram(bins = 60,
                 col = "white",
                 fill = "#9099A2"
                 ) +
  theme_tufte() +
  labs(
    title = "Shared receipt with POI"
    ) +
  labs(
    x = "Number log10",
    y = "Count"
    ) + scale_x_log10()

grid.arrange(plt20, plt21, plt22, plt23, plt24, ncol = 3)
```

The data transfromations give a better visualisation of the distribution
of values, a number of which have more of a normal distribution after a log10
transformation.

```{r echo=FALSE, BPlots31, fig.height = 6, fig.width = 4, fig.align="center"}
freq_poly <-  function(x_v, name, scale, 
                       dataset=enron, group=enron$poi) {
plt <- qplot(x = x_v, data = dataset,
             geom = 'freqpoly', 
             color = group,
             bins = 25) +
  scale_x_continuous() +
  theme_tufte() +
  labs(
    title = name
    ) +
  labs(
    x = scale
      )
plt + scale_color_manual(values = c('#0B3C5D', '#D9B310'))
}

pltf6 <- freq_poly(enron$bonus, 
  "Bonus", "$ log 10")
pltf6 <- pltf6 + scale_x_log10()
pltf7 <- freq_poly(enron$deferral_payments, 
  "Deferral payments", "$ log 10")
pltf7 <- pltf7 + scale_x_log10()
pltf8 <- freq_poly(enron$deferred_income, 
  "Deferred income", "$")

pltf10 <- freq_poly(enron$exercised_stock_options, "Exercised stock options", "$")
pltf11 <- freq_poly(enron$expenses, "Expenses", "$")
pltf12 <- freq_poly(enron$loan_advances, "Loan advances", "$")
pltf13 <- freq_poly(enron$long_term_incentive, "Long term incentive", "$")
pltf14 <- freq_poly(enron$other, "Other", "$")
pltf15 <- freq_poly(enron$restricted_stock, "Restricted stock", "$")
pltf16 <- freq_poly(enron$restricted_stock_deferred, "Restricted stock deffered", "$")
pltf17 <- freq_poly(enron$salary, "Salary", "$")
pltf18 <- freq_poly(enron$total_payments, "Total payments", "$")
pltf19 <- freq_poly(enron$total_stock_value, "Total stock value", "$")

multiplot(pltf6, pltf7, pltf8, pltf10, 
          pltf11, pltf12, pltf13, pltf14, pltf15, 
          pltf16, pltf17, pltf18, pltf19, cols = 3)
```

For the financial variables frequency polygons are used to investigate persons
of interest.

A few of the variables are difficult to seperate any trends between POI and
normal people. 

```{r echo=FALSE, BPlots1, fig.height = 3, fig.width = 4, fig.align="center"}
pltf5 <- freq_poly(enron$shared_receipt_with_poi, 
  "Shared receipt with POI", "Number log 10")
pltf5 <- pltf5 + scale_x_log10()
pltf1 <- freq_poly(enron$from_messages, 
  "From messages", "Number log 10")
pltf1 <- pltf1 + scale_x_log10()
pltf2 <- freq_poly(enron$from_this_person_to_poi, 
  "From this person to POI", "Number log 10")
pltf2 <- pltf2 + scale_x_log10()
pltf3 <- freq_poly(enron$to_messages, 
  "To messages", "Number log 10")
pltf3 <- pltf3 + scale_x_log10()
pltf4 <- freq_poly(enron$from_poi_to_this_person, 
  "From POI to this person", "Number log 10")
pltf4 <- pltf4 + scale_x_log10()

multiplot(pltf1, pltf2, pltf3, pltf4, pltf5, cols = 3)
```

Looking at POI within the email data highlights some promising variables.
Shared receipt with POI has a spike for true POI but is overlain by a number
of non-POI responses as well.

From POI to this person suggests higher numbers can be related to other POIs.

From this person to POI is challenging, with mixed occurences of True within
the distribution.

```{r echo=FALSE, freq_pair_plot, fig.height = 9, fig.width = 9, message=FALSE, warning=FALSE,}
drops <- c("X","name", "email_address", "loan_advances")

ggpairs(enron[ , !(names(enron) %in% drops)],
        title = "Enron Dataset Pair Plot"
        ) +
  theme_minimal()
```

The pair plot gives a way to quickly see any highly correlated variables.

Loan advances is removed as it has too few data points.

```{r echo=FALSE, scatter, fig.height = 5, fig.width = 5}
ggplot(aes(x = enron$shared_receipt_with_poi, y = enron$to_messages),
       data = enron) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Density vs Fixed Acidity",
    x = "Shared receipt with poi",
    y = "To messages"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

The

```{r echo=FALSE, scatter2, fig.height = 5, fig.width = 5}
ggplot(aes(x = enron$from_messages, y = enron$from_this_person_to_poi),
       data = enron) +
  geom_point(alpha = 0.2,
             position = position_jitter(),
             color = '#192231') +
  theme_tufte() +
  labs(
    title = "Density vs Fixed Acidity",
    x = "from message",
    y = "from this person to poi"
    ) +
  geom_smooth(method = 'lm', formula = y~x,
              colour = '#192231')
```

The

```{r echo=FALSE, Multivariate_Plots_7, fig.height = 6, fig.width = 9}
plt_mv7 <- ggplot(aes(x = log10(enron$from_this_person_to_poi/enron$from_messages),
                  y = log10(enron$from_poi_to_this_person/enron$to_messages)),
       data = enron) +
  geom_point(alpha = 0.01,
             position = position_jitter()
             ) +
  theme_tufte() +
  labs(
    title = "Attempting to seperate persons of interest in email data",
    x = "from messages/from this person to poi",
    y = "to messages/from poi to this person"
    )
fnl_plt_4 <- plt_mv7 + geom_point(aes(size = enron$shared_receipt_with_poi), 
                                  pch = 21, 
                                  show.legend = T,
                                  alpha = 0.5) + 
  aes(fill = factor(poi)) + geom_text(
    aes(label = name),
    hjust = 0, 
    nudge_x = 0.03,
    colour = "#9099A2",
    check_overlap = TRUE,
    size = 3
    )
fnl_plt_4
```

The plot uses ratios of email variables to highlight the persons of interest
and how they vary from non-persons of interest.

```{r echo=FALSE, Multivariate_Plots_8, fig.height = 6, fig.width = 9}
plt_mv7 <- ggplot(aes(x = log10(enron$total_payments/enron$bonus),
                  y = (enron$salary)),
       data = enron) +
  geom_point(alpha = 0.01,
             position = position_jitter()
             ) +
  theme_tufte() +
  labs(
    title = "Attempting to seperate persons of interest in financial data",
    x = "log10 total payments / bonus",
    y = "salary $"
    )
fnl_plt_4 <- plt_mv7 + geom_point(aes(size = sqrt(enron$exercised_stock_options)), 
                                  pch = 21, 
                                  show.legend = T,
                                  alpha = 0.5) + 
  aes(fill = factor(poi)) + geom_text(
    aes(label = name),
    hjust = 0, 
    nudge_x = 0.03,
    colour = "#9099A2",
    check_overlap = TRUE,
    size = 3
    )
fnl_plt_4
```

The

```{r echo=FALSE, Multivariate_Plots_8, fig.height = 6, fig.width = 9}
plt_mv7 <- ggplot(aes(x = log10(enron$/enron$bonus),
                  y = (enron$salary)),
       data = enron) +
  geom_point(alpha = 0.01,
             position = position_jitter()
             ) +
  theme_tufte() +
  labs(
    title = "Attempting to seperate persons of interest in financial data",
    x = "log10 total payments / bonus",
    y = "salary $"
    )
fnl_plt_4 <- plt_mv7 + geom_point(aes(size = sqrt(enron$exercised_stock_options)), 
                                  pch = 21, 
                                  show.legend = T,
                                  alpha = 0.5) + 
  aes(fill = factor(poi)) + geom_text(
    aes(label = name),
    hjust = 0, 
    nudge_x = 0.03,
    colour = "#9099A2",
    check_overlap = TRUE,
    size = 3
    )
fnl_plt_4
```